The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2019-10-18 categorical-zero-one.tex
Rewrite sections on infinite tensor products

\documentclass[11pt]{article}
\usepackage{eigilscmds}
\usepackage{tikzit}
\usepackage{geometry}
\input{comonoids.tikzdefs}
\input{comonoids.tikzstyles}

@@

Of course, in a general Markov category the above cofiltered limit need not exist for infinite $J$, in which case the infinite tensor product does not exist either (as in \Cref{infprods_finstoch}).

\begin{remark}
	\label{double_infinite_product}
    It is easily verified that 
    \[\bigotimes_{i \in J} X_i \tensor \bigotimes_{i' \in J'} X_{i'} \cong \bigotimes_{i \in J \sqcup J'} X_i\]
    i.e tensor products of infinite tensor products can be written as single infinite tensor products in the obvious way.
\end{remark}

\todo[inline]{Probably the similar theorem about infinite tensor products of infinite tensor products is true as well, although I have not verified it.}

\section{Infinite tensor products in Markov categories}

@@

\begin{definition}
	\label{defn_kolmogorov_ext}
	A Markov category \emph{satisfies Kolmogorov extension} if it has infinite tensor products $\bigotimes_{i \in J} X_i$ such that 
	\[
		\tikzfig{infinite_product_copy}
	\]
\end{definition}

@@

The idea is that this equation uniquely determines the comonoid structure: by \Cref{double_infinite_product}, the comultiplication
\[
	\mathsf{copy}_{\bigotimes_i X_i} \: : \: \bigotimes_i X_i \longrightarrow \left( \bigotimes_i X_i \right) \otimes \left( \bigotimes_i X_i \right)
\]
is uniquely determined by its composition with the maps $\pi_i \otimes \pi_j$ for all $i,j \in J$.

\todo[inline]{Can we show that the comonoid laws automatically follow from \Cref{defn_kolmogorov_ext}?}

In particular, the $i=j$ case of the equation above shows that the marginalizations $\pi_i$ are deterministic.

@@

But in case that the subcategory of deterministic maps $\cC_{\det}$ has infinite products in addition to finite ones, it seems natural to ask whether this is compatible with the structure of $\cC$ in a suitable sense.

\begin{lemma}
	If $\cC$ satisfies Kolmogorov extension, then an infinite tensor product $\bigotimes_i X_i$ is a product $\prod_i X_i$ with projections $\pi_i$ in the subcategory of deterministic morphisms $\cC_\detc \subseteq \cC$.
\end{lemma}

\begin{proof}
	\todo[inline]{We probably have to show that all finite marginalization maps $\pi_F$ are deterministic}
\end{proof}

\begin{remark}
    If we take as input certain maps $I \to X_i$ as well, so that we have an infinite system of probability distributions, we can also try to define the infinite tensor product as a colimit, which will then be a space with a probability distribution, corresponding to an infinite system of random variables.
    It would also be possible to formulate the zero-to-one law in this setting, but we have not explored this.
\end{remark}

@@

This argument also shows that this Markov kernel is determined uniquely by the fact that $p(a,\pi_F\inv(B)) = p_F(a,B) = \pi_F \circ p$.
This proves that $(X,\Sigma)$ is the limit of $(X_F,\Sigma_F)$, so that it is the tensor product $\bigotimes_{i\in J} X_i$.
\end{example}

\begin{example}
	\label{infprods_finstoch}
	$\sf{FinStoch}$ does not have infinite tensor products.
\end{example}

@@

\begin{corollary}
    Let $\{X_i\}_{i \in J}$ be a family of topological spaces, $Y$ a Hausdorff space, and let $f: \prod_i X_i \to Y$ be a continuous function
    which is independent of any finite prefix of the input.
    Then the map $f$ is constant.
\end{corollary}
\begin{corollary}
    Let $X, Y$ be topological spaces, $Y$ Hausdorff, and let $f: \prod_{i\in J} X \to Y$ be a continuous function which is independent of any finite permutation of the input.
    Then $f$ is constant.
\end{corollary}
\begin{proof}
    The theorems follow from \cref{thm:kolmog} and \cref{thm:hevsav}, applied to the Kleisli category $Kl(H)$ of the hyperspace monad,
    using the maps $* \to X_i$ and $* \to X$ given by the closed subsets $X_i \subseteq X_i, X \subset X$.
    The conclusion is that the closure of the image, call it $A$, has the following property: $\overline{\{(a,a) \mid a \in A\}} = A \times A$ as subsets of $Y \times Y$.
    But since the diagonal is closed in $Y$, this clearly implies that $A$ must be a singleton.
\end{proof}

>>>>

\documentclass[11pt]{article}
\usepackage{eigilscmds}
\usepackage{tikzit}
\usepackage{geometry}
\usepackage{enumerate}
\input{comonoids.tikzdefs}
\input{comonoids.tikzstyles}

@@

Of course, in a general Markov category the above cofiltered limit need not exist for infinite $J$, in which case the infinite tensor product does not exist either (as in \Cref{infprods_finstoch}).
\section{Infinite tensor products in Markov categories}

@@

\begin{definition}
    \label{defn_kolmogorov_ext}
    Let $\cC$ be a Markov category, let $\{X_i\}_{i\in J}$ be an infinite family of objects, and let $\bigotimes_i X_i$ be an infinite tensor product.
    We say this product \emph{satisfies Kolmogorov extension} if the following two conditions hold:
    \begin{enumerate}[(1)]
        \item The finite marginals \[\pi_F: \bigotimes_{i\in J} X_i \to \bigotimes_{i \in F} X_i\] are deterministic
        \item These maps exhibit $\bigotimes_{i\in J} X_i$ as the cofiltered limit of the diagram $F \mapsto \bigotimes_{i \in F} X_i$ \emph{in $\cC_{\rm det}$}.
    \end{enumerate}
\end{definition}

@@

\begin{remark}
    Since the (finite) tensor product is the product in $\cC_{\rm det}$, condition (2) is equivalent to the statement that the marginals
    \[\pi_j: \bigotimes_i X_i \to X_j\]
    exhibit $\bigotimes_i X_i$ as the product of the $X_j$ in $\cC_{\rm det}$,
    since in any category, \[\lim_{F \subset J} \prod_{i \in F}X_i \cong \prod_{i \in J}X_i\]
\end{remark}

\begin{remark}
    Condition (2) is also equivalent to the statement that, given a compatible family of deterministic maps $a_F: A \to \bigotimes_{i \in F}$, the induced map $a: A \to \bigotimes_i X_i$ is deterministic.
\end{remark}

This determines the comonoid structure uniquely: $\mathsf{copy}_{\bigotimes_i X_i}$ has to be the diagonal map in $\cC_{\rm det}$,
and this condition determines $\bigotimes_i X_i$ up to \emph{deterministic} isomorphism.

@@

\begin{lemma}
    Let $J,J'$ be possibly infinite sets, and $\{X_i\}_{i\in J}, \{Y_i\}_{i\in J'}$ two families of objects in $\cC$, so that both have infinite tensor products with Kolmogorov extension.
    Let $\{Z_i\}_{i\in J \sqcup J'}$ be a family of objects indexed by the disjoint union, where $Z_i = X_i$ for $i\in J$, $Z_i = Y_i$ for $i \in J'$,
    and suppose $\{Z_i\}$ admits an infinite tensor product with Kolmogorov extension, too.
    Consider the family of maps
    \[\rho_F: \bigotimes_{i\in J}X_i \tensor \bigotimes_{i\in J'} Y_i \labelto{\pi_{F \cap J} \tensor \pi_{F \cap J'}} \bigotimes_{i \in F \cap J}X_i \tensor \bigotimes_{i\in F \cap J'}Y_i\]
    indexed by finite subsets $F \subset J \sqcup J'$ of the disjoint union.
    The induced map
    \[\bigotimes_{i\in J}X_i \tensor \bigotimes_{i\in J'} Y_i \to \bigotimes_{i \in J \sqcup J'} Z_i\]
    is a deterministic isomorphism.
\end{lemma}
\begin{proof}
    It is clear that the $\rho_F$ are deterministic, and that they form a limit diagram in $\cC_{\rm det}$.
    Hence the induced map into the actual limit in $\cC_{\rm det}$ is an isomorphism.
\end{proof}

@@

This argument also shows that this Markov kernel is determined uniquely by the fact that $p(a,\pi_F\inv(B)) = p_F(a,B) = \pi_F \circ p$.
This proves that $(X,\Sigma)$ is the limit of $(X_F,\Sigma_F)$, so that it is the tensor product $\bigotimes_{i\in J} X_i$.
It is simple to verify that it also has Kolmogorov extension, using e.g. \cite[Example~10.4]{markov_cats}
\end{example}

\begin{example}
	\label{infprods_finstoch}
	$\sf{FinStoch}$ does not have infinite tensor products.
\end{example}
\todo[inline]{Add an example where $\cC_{\rm det}$ has infinite products, but $\cC$ lacks infinite tensor products? Like sets and multivalued functions?}

From this point, the term \emph{infinite tensor product} will refer always to an infinite tensor product with Kolmogorov extension.

@@

\begin{corollary}
    Let $\{X_i\}_{i \in J}$ be a family of topological spaces, $Y$ a Hausdorff space, and let $f: \prod_i X_i \to Y$ be a continuous function
    which is independent of any finite prefix of the input.
    Then the map $f$ is constant.
\end{corollary}
\begin{corollary}
    Let $X, Y$ be topological spaces, $Y$ Hausdorff, and let $f: \prod_{i\in J} X \to Y$ be a continuous function which is independent of any finite permutation of the input.
    Then $f$ is constant.
\end{corollary}
\begin{proof}
    The theorems follow from \cref{thm:kolmog} and \cref{thm:hewsav}, applied to the Kleisli category $Kl(H)$ of the hyperspace monad,
    using the maps $* \to X_i$ and $* \to X$ given by the closed subsets $X_i \subseteq X_i, X \subset X$.
    The conclusion is that the closure of the image, call it $A$, has the following property: $\overline{\{(a,a) \mid a \in A\}} = A \times A$ as subsets of $Y \times Y$.
    But since the diagonal is closed in $Y$, this clearly implies that $A$ must be a singleton.
\end{proof}
